### 2025-09-22 paper/sec/introduction.tex
Start writeup of table 1.

NLP aspect of the problem is potentially a big problem space in itself. We will restrict interest to certain
idiomatic uses of natural language in making/justifying scientific claims. Table~\ref{tab:fluid_examples}

>>>>

Table~\ref{tab:fluid_examples} summarises the natural language idioms we have studied in this paper. In the
realm of traditional quantitative expressions, and when using state-of-the-art models like gpt4o and gpt5, our
system is able to recognise direct numerical values, percentages, averages, minima and maxima, ranks, and
totals, each mapped to a corresponding computation over the source dataset. For example, phrases such as ``the
Energy Sector accounts for 52.80\% of total emissions'' or ``the expected projected methane emissions for 2030
is 13.5'' are interpreted as division and aggregation operations over data values. Similarly, ``recorded its
highest emissions in 2030'' is interpreted as a `maximumBy` query, while rank-based statements such as ``CNN
gives the lowest accuracy'' are interpreted as ranking computations.

We also consider \emph{trend} expressions, which include relative or comparative natural language phrases
describing how some property evolves over time. These include single comparisons (e.g.~``training time growing
from 67 to 106 seconds'') and pairwise comparisons (e.g. “SVM outperformed all other kernels, offering better
precision”). Such idioms are mapped to higher-level operators like trendWord, growShrink, and betterWorse,
which combine statistical aggregation with semantic markers of relative performance. Taken together, these
categories cover a substantial portion of the numerical reasoning idioms found in empirical scientific
writing.

Notably, the current coverage omits several common linguistic forms. We have yet to study the ability of LLMs
to interpret approximate quantitative terms like ``around 50\%'' or ``roughly 100 instances'', nor
interval-based descriptions such as ``between 30 and 40\%” or ``within 5–10 seconds''. We have no reason for
thinking these will present specific difficulties. ``Graded'' modal adjectives like ``slightly better'',
``significantly higher'' present more of a challenge linguistically~\cite{lassiter17}. Likewise, idioms
expressing trends over time in nuanced or qualified ways (e.g. ``steady increase'', ``sharp decline'') or
distributional descriptors like ``most models'', ``the majority of cases'', or ``usually performs better
than'' present significant challenges to formal interpretation because of their reliance on non-standard
quantifiers. Extending coverage to these idioms would substantially deepen our tool's ability to bridge
natural language reporting with interpretation in terms of the underlying dataset. We discuss this further in
\secref{future-work}.
